---
title: "Toy example with ggnetworkx"
author: "CSL"
date: "2019-11-5"
output:
  html_document:
    keep_md: true
---

[genenetworkx](https://github.com/KlausVigo/ggnetworx)

To install:
```{r, eval=FALSE}
devtools::install_github("GuangchuangYu/ggtree")
devtools::install_github("KlausVigo/ggnetworx")
install.packages("rgl")
```


```{r}
library(ggnetworx)
?ggnetworkx  ## we get the examples
```

There are two options: 1) split networks, and 2) explicit networks (from extended newick).

## Split networks

```{r}
### split networks
data(yeast, package="phangorn")
yeast ## sequences
dm <- phangorn::dist.ml(yeast) 
dm  ## distance matrix
nnet <- phangorn::neighborNet(dm)
nnet
ggnetworx(nnet) + geom_tiplab2()
plot(nnet)
```

## Explicit networks

```{r}
### explicit networks
enet <- ape::read.evonet(text="((a:2,(b:1)#H1:1):1,(#H1,c:1):2);")
enet  ## explicit network
ggevonet(enet) + geom_tiplab() ## from ggnetworkx
plot(enet)  ## from ape
```


Now, playing around, how do branch lengths affect:
```{r}
### just adding branch length to minor edge (#H1,c)
enet1 <- ape::read.evonet(text="((a:2,(b:1)#H1:1):1,(#H1:1,c:1):2);")
ggevonet(enet1) + geom_tiplab() ##nothing happens

## apparently, you do need ultrametric networks, bc the following does not work
enet2 <- ape::read.evonet(text="((a:0.22,(b:0.4)#H1:0.3):0.12,(#H1,c:0.15):0.2);")
ggevonet(enet2) + geom_tiplab()
plot(enet2) ## but with ape it works
```

Maybe not entirely ultrametric, but at least the donor and recipient of gene flow should have similar length:
```{r}
enet3 <- ape::read.evonet(text="((a:0.22,(b:0.4)#H1:0.3):0.12,(#H1,c:0.45):0.2);")
ggevonet(enet3) + geom_tiplab()
plot(enet3)
plot(enet3, type="c")
```

## ggtree

There is an option in `ggtree` to plot as cladogram: `branch.length="none"`, can we try?

```{r}
ggevonet(enet3,branch.length="none") + geom_tiplab()
```

But still does not work with the weird branches. The time starts at the root, and on one side there is `0.12+0.3=0.42`, but on the other side, there is `0.2+0.15=0.35`, so the tip `c` is drawn before the hybrid edge.
I believe that ultrametric should not be the default.

```{r}
ggevonet(enet2,branch.length="none") + geom_tiplab()
```

```{r}
ggevonet(enet2,layout="slanted") + geom_tiplab()
```

Thus, we still need to figure out why this network is weird.

### Does it allow inheritance probs?

Yes, it does not break, but it does not matter if they don't add up to one:
```{r}
enet4 <- ape::read.evonet(text="((a:2,(b:1)#H1:1::0.99):1,(#H1:1::0.2,c:1):2);")
ggevonet(enet4) + geom_tiplab()
```
## Trying with bigger networks

```{r}
wnet <- ape::read.evonet(text="(Mno00,Bph03,(((Bmu,(Bbo01,Bbo02):9.41949690722463):1.6484045144986545,(Bac00,((balMys,Egl00):9.055782863234407)#H13:5.268095830633961::0.8331903627712247):1.100383484076166):0.066043783465627,((Erob02,Erob01):9.997600013276905,#H13:0.19346605673125924::0.16680963722877526):0.35164828246609625):2.146642742731017);")
ggevonet(wnet) + geom_tiplab()
```

We get an error/warning that some NA values in `edge.length`.

We can compare to ape:
```{r}
plot(wnet)
```

The plot from ape looks weird because hybridization starts on internal nodes.

Now, if we try the `slanted` option, the hybrid edge still looks horizontal. This is different behavior to the plot of `enet2`.
```{r}
ggevonet(wnet, layout="slanted") + geom_tiplab()
```

### Now more than one hybridization:

```{r}
fishnet <- ape::read.evonet(text="(Xalvarezi,Xmayae,((Xsignum,((Xmonticolus,(Xclemenciae_F2,#H25:::0.032485577051040423):0.979116879342896):1.4858786525152954,(((((((((Xgordoni,Xmeyeri):0.26111415598145865,Xcouchianus):3.5325554148283564,Xvariatus):0.6445452395880061,Xevelynae):0.41219807813960146,(Xxiphidium,#H24:0.0::0.16538333427567822):1.4020042304402076):0.2921320715527958,Xmilleri):0.4688709909949548,Xandersi):0.6492623942884177,Xmaculatus):1.0200542801748702,(((Xmontezumae,(Xcortezi,(Xbirchmanni_GARC,Xmalinche_CHIC2):0.8943004029517725):0.6903622798656042):0.4926275928330048,((Xnigrensis,Xmultilineatus):1.4508210422790195,(Xpygmaeus,Xcontinens):1.9686498919562234):2.428065080719157):2.0435103751744865)#H24:0.6476450483433508::0.8346166657243218):0.7816554661308311):1.9841381634366537):0.31614675520900004,(Xhellerii)#H25:::0.9675144229489596):0.28226883225049243);")
ggevonet(fishnet) + geom_tiplab()
plot(fishnet)
```

Black hybrid edges look confusing as they appear as polytomies.

We can compare in the backgroung with julia plot:
```
using PhyloNetworks
using PhyloPlots # to visualize networks
using RCall      # to send additional commands to R like this: R"..."

net = readTopology("(Xalvarezi,Xmayae,((Xsignum,((Xmonticolus,(Xclemenciae_F2,#H25:::0.032485577051040423):0.979116879342896):1.4858786525152954,(((((((((Xgordoni,Xmeyeri):0.26111415598145865,Xcouchianus):3.5325554148283564,Xvariatus):0.6445452395880061,Xevelynae):0.41219807813960146,(Xxiphidium,#H24:0.0::0.16538333427567822):1.4020042304402076):0.2921320715527958,Xmilleri):0.4688709909949548,Xandersi):0.6492623942884177,Xmaculatus):1.0200542801748702,(((Xmontezumae,(Xcortezi,(Xbirchmanni_GARC,Xmalinche_CHIC2):0.8943004029517725):0.6903622798656042):0.4926275928330048,((Xnigrensis,Xmultilineatus):1.4508210422790195,(Xpygmaeus,Xcontinens):1.9686498919562234):2.428065080719157):2.0435103751744865)#H24:0.6476450483433508::0.8346166657243218):0.7816554661308311):1.9841381634366537):0.31614675520900004,(Xhellerii)#H25:::0.9675144229489596):0.28226883225049243);")

R"name = function(x) file.path('.', x)" # function to create file name in appropriate folder
R"pdf(name('snaqplot.pdf'), width=4, height=3)"
R"par"(mar=[0,0,0,0]) # to reduce margins (no margins at all here)
plot(net, :R, showGamma=true); # network is plotted & sent to file
R"dev.off()"; # wrap up and save image file

knitr::include_graphics("snaqplot.pdf")
```

